### Course - Reproducible Research
#### Week 2 Assignment - R Markdown & Knitr
Suleman Wadur 


Loading data file 
```{r Load_data, echo=TRUE}
## Turns off exponential notation of numeric values such as when using the mean function
options(scipen = 999)

## Load needed libraries
library(dplyr)


## Set working directory
workdir <- "C:/Move 4/Coursera/DataScience/Course5-ReproducibleResearch/Week2/Assignment/RepData_PeerAssessment1"
setwd(workdir)



if (file.exists("activity.zip") && !file.exists("activity.csv")) {
  unzip(zipfile="./activity.zip", exdir=".")
}

if(!file.exists("activity.csv")) {
  stop("One of the input files is missing!")
}

activitydata <- read.csv("activity.csv")
  
```
  
  
  
#### Mean total number of steps taken per day
```{r TotalSteps}
#plot(activitydata$date, mean(is.na(activitydata$steps)))  DailyMean=activitydata$steps
data0 <- subset(activitydata, !is.na(steps))
TotalStepsByDay <- setNames(aggregate(data0$steps, list(data0$date), FUN=sum), c("Day", "Total.Steps"))
#plot(data1$x, data1$Day, type = "h", xlab = "Day", ylab = "Mean daily steps", main = "Average total steps per day")

TotalStepsByDay
```
  
Histogram: total number of steps taken per day
```{r HistogramTotalSteps}
hist(TotalStepsByDay$Total.Steps, main = "Histogram: Total Number of Steps per Day", col = "lightblue", xlab = "Total Steps per day" )
```

Mean and Median of Steps Taken per day
```{r MeanMediaSteps}
meanOfSteps <- mean(TotalStepsByDay$Total.Steps)
medianOfSteps <- median(TotalStepsByDay$Total.Steps)
```
Mean is `r meanOfSteps` and median is `r medianOfSteps`  
  
  
#### Time Series: Average Daily Activity Pattern
```{r AverageDailyActivityPattern}
ActivtyPattern <- aggregate(steps ~ interval, data0, mean)
plot(ActivtyPattern$interval, ActivtyPattern$steps, type = "l", main="Average Daily Activity Pattern", ylab = "Average Daily Steps", xlab = "5-minute Interval")

maxSteps <- ActivtyPattern[which.max(ActivtyPattern$steps),]

```
5-minute interval with the maximum number of steps is `r maxSteps$interval` with `r maxSteps$steps` steps



#### Imputing missing values
```{r MIssingValues}
# get a vector of complete cases. This will return true or false, with rows having missing values as TRUE
y <- !complete.cases(activitydata)

# counts the number of records with TRUE, indicating they have missing records.
mrecords <- sum(y)
```
There are a total of `r mrecords` records with missing values
  
  
Using the mean of 5-minute interval, a new variable is added for records with missing steps...
```{r newData}
#create a new variable with the copy of the steps data. This will be updated in the function below
activitydata$new.steps <- activitydata$steps

#determine average steps by interval
meansByInterval <- aggregate(steps ~ interval, data = activitydata, mean)

# gets index of all rows in the original dataset with missing data
p <- which(is.na(activitydata), arr.ind=TRUE)

for(i in 1:nrow(p)) {
  j <- p[i,1]
  
  #gets index of the mean interval stored in meansByInterval data frame
  indexOfMeanInterval = which(meansByInterval$interval == activitydata[i,3])
  
  #update the new steps with the mean of its interval
  activitydata[j, 4] <- meansByInterval[indexOfMeanInterval,2]
}
```


Calculating summary on updated data...
```{r HistogramStepsUpdated}
TotalStepsByDay <- setNames(aggregate(activitydata$new.steps, list(activitydata$date), FUN=sum), c("Day", "Total.Steps"))
hist(TotalStepsByDay$Total.Steps, main = "Histogram: Total Number of Steps per Day: Missing Data corrected", col = "lightblue", xlab = "Total Steps per day" )
```

```{r MeanMediaStepsUpdated}
NewMeanOfSteps <- mean(TotalStepsByDay$Total.steps)
NewMedianOfSteps <- median(TotalStepsByDay$Total.steps)
```
Old Mean: `r meanOfSteps` and Old median: `r medianOfSteps`  
New Mean: `r NewMeanOfSteps` and New median: `r NewMedianOfSteps`  

From the histogram above, the frequency has certainly increased. Only the median total daily steps seems to have changed when the missing data got updated.  


#### Activity Pattern: Weekdays vs Weekends
```{r WeekDayPattern, fig.height=10}
#loop through data and derive the weekend/weekday
for (i in 1:nrow(activitydata)) {
  if (weekdays(as.Date(activitydata$date[i])) == "Saturday" | weekdays(as.Date(activitydata$date[i])) == "Sunday")
      activitydata$weekday[i] <- "weekend"
  else
    activitydata$weekday[i] <- "weekday"
}

TotalStepsByWeekDay <- setNames(aggregate(activitydata$new.steps~activitydata$weekday+activitydata$interval, list(activitydata$weekday), FUN=mean), c("Weekday", "Interval", "Average.Steps"))

#TotalStepsByWeekDay


#defines a 2 row, 1 column panel plot area
par(mfrow = c(2,1) )

#subset data according to weekend or weekday and plot the time series graph
dday <- subset(TotalStepsByWeekDay, Weekday == "weekday")
dend <- subset(TotalStepsByWeekDay, Weekday == "weekend")
plot(dday$Interval, dday$Average.Steps, type = "l", main="Activity Pattern - Weekday", ylab = "Average number of Steps", xlab = "5-minute Interval", col = "lightblue" )

plot(dend$Interval, dend$Average.Steps, type = "l", main="Activity Pattern - weekend", ylab = "Average number of Steps", xlab = "5-minute Interval", col = "lightblue")
```

